
\GenericNonNumberedSection{Acknowledgements}{
Graduate studies is a gargantuan challenge full of difficulties, disappointment
and discouragement. It is also greatly enriching, satisfying and something to
be proud of. Even though I present here my own thesis, many great people were
behind me and helping me surpass myself.

I would like to first thank my supervisor, professor Lora Ramunno, for allowing
me to pursue my graduate studies under her umbrella. I am glad she trusted me
as her first PhD student and I hope to have come up to her expectations. Even
though difficulties sometimes emerged, I am greatly thankful of the trust she
had in me. She gave me the freedom and the tools I needed to fulfill my goals
and ideas. She gave me the responsibility of both the purchase and maintenance
of her two clusters, an invaluable experience. She advised me wisely during the
years. I hope the end of my studies will not mean the end of some collaboration.

I would also like to thank Edward ``Eddie'' Ackad for the numerous and deep
discussions on both professional and personal mater. He never refused to answer
my questions, even pushing me to answer them myself instead of blindly answering.
You are a good friend that I'm already missing.

Moving to a different city, in a different culture and language can be
challenging, but the many friends I've made in Ottawa supported me without
questions. I must name Charles Varin and Martin Bertrand who became true
friends that I will hopefully keep after my move back to Montreal.

I am grateful to my family who encouraged me before and during those years,
supporting me into never giving up -- \textit{encore un p'ti coup de coeur!}
My parents, brother, sisters who helped me moved in during this emotional time.
I want to show gratitude to my in-laws for supporting me without question
and understanding my countless absences; the sentence \textit{tonton travail?}
will simply be a memory from now on.

I cannot thank enough my beautiful and sweet Christine that supported me during
all these long years, remotely at first and more closely during the last one.
You've been so supportive, patient and understanding with me, at all times,
the best as the worsts. I know it's been hard for you and I deeply apologize
for everything I made you went through; listening to my grumbles,
seeing me leave every Sunday night. You never stopped believing in me.
I deeply thank you for always being there for me. You said I was your pillar
during your own PhD. I'm saying you are my keystone, keeping me in balance,
preventing me from falling, strengthening me. You've been a saint to endure my
countless absent hours. I sincerely cannot wait to be back in your arms and
enjoy our time together -- as a family of soon three. I cannot wait to
see our \textit{petit poulet} and show him the world.

I love you, and always will.
}
