% \subsection{Detailed Introduction}
% Describe in more details the work done: The physics, the models. The specific
% litterature search goes here.

The previous section described the general goals of this thesis and was used to
give the reader an overview picture of the field. The following will go
into more details.

NOTES:
\begin{itemize}
\item Current work motivated by recent FEL experiments at specific regime.
\end{itemize}


\subsection{Goals}
In view of the previous section, the goal of the current thesis is to increase
to global knowledge of laser-matter interaction. More specifically, the
question of \textit{how} the energy is deposited in a nanoscale object by an
ultra-short and ultra-intense laser pulse is important.

\cite{Young2010,Chapman2011}

\subsection{Laser-Cluster interaction}
\cite{Reinhard2004}


\subsubsection{Clusters}
\label{section:intro:clusters:clusters}

Clusters of atoms are an extremely useful tool to study laser-matter
interaction. They are experimentally easy to produce since many
years\cite{Haberland1994}. Their size and constitution can be varied almost at
will \cite{Martin1996}.

\cite{Fennel2010}

\subsubsection{Heating mechanisms}
\label{section:intro:clusters:heating}
FIXME: Rename ``heating'' to something else.

Specify that the different mechanisms depends on the regime (material and
wavelength).

Don't talk about tunnel ionization. Just mention it, maybe one paragraph (not
one page). Giving Keldysh parameter, we are most interested in another regime.


\subsubsection{Studied parameters}
NOTES:
\begin{itemize}
\item Put here why we chose a classical model. Cite
Reinhard2004\cite{Reinhard2004}.
\end{itemize}




\input{introduction_tools.tex}

\subsection{Thesis Outline}
Don't forget a review/description of each sections of the thesis.

Include a subsection with each of my contributions from each paper.

\ReferencesSubsection{references_intro}
